{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-12-05 ident_manu.Rnw
first two pages of discussion

\begin{abstract}
\noindent Bio-geo-chemical models are useful tools in environmental sciences that can guide management and policy-making. Consequently, significant time and resources are spent developing these models in system-specific contexts. The optimization of model parameters to maximize precision, including transferability of these models to different systems, are fundamental concerns in the development and application of these tools. This study provides a context for understanding quantitative limitations of coupled hydrodynamic-ecological models by evaluating parameter sensitivity and identifability of a \ac{zerod} unit of a larger spatio-temporal model of hypoxia on the \acl{lcs} of Gulf of Mexico.  The analysis provides a contrast of numeric and ecological certainty in parameter subsets using a systematic framework to infer larger trends in dissolved oxygen dynamics over time, having implications for understanding factors that contribute to environmental conditions that are detrimental to aquatic resources.  In particular, we focus on issues of parameter identifiability using local sensitivity analyses to provide quantitative descriptions of numerical constraints on model precision.  We argue that quantitative and ecological certainty in model calibration are often at odds and practitioners must choose explicit model components to optimize given tradeoffs between the two. We further conclude that numerically optimal parameter sets for models of hypoxia are often small subsets of the complete parameter set because of redundancies in the unique effects of paramater perturbations on model output.  As a result, we demonstrate that use of a model for inference into ecological mechanisms of observed or predicted changes in hypoxic condition can be potentially misguided in the absence of quantitative descriptions of identifiability.  Although these concerns have been expressed in the literature, they are rarely explicitly addressed or included in model evaluations.  In addition to immediate implications for regional models, we provide a framework for describing the effects of parameter uncertainty and identifiability that can be applied to similar models to better inform environmental management.
\end{abstract}
\acresetall

@@

Hypoxia formation in bottom waters of coastal oceans occurs primarily from excess nutrient inputs from land-based sources \citep{Justic87,Diaz95,Howarth96}.  These events are detrimental to aquatic organisms and have significant negative effects on economic resources derived from coastal ecosystems \citep{Lipton03,Diaz11}.  An understanding of the biological, physical, and chemical processes that contribute to the growth of hypoxic areas is a critical concern for mitigating and preventing these negative impacts.  Numerical ecosystem models have been important tools for describing current knowledge of ecosystem processes that contribute to hypoxia formation and for predicting the effects of proposed management activities or future scenarios \citep{Scavia04,Hagy07,Pauer16}.  Unlike statistical models that have more generic structures, simulation and process-based models include explicit descriptions of relevant processes that are constrained by empirical or observational data relevant to the system of interest \citep[e.g.][]{Omlin01b,Eldridge10}.  These models are often coupled with hydrodynamic grids to provide spatially-explicit representations of patterns in three dimensions \citep{Warner05,Zhao10,Ganju16}. Combined hydrodynamic and bio-geo-chemical models have been developed specifically to describe hypoxic conditions on the \ac{lcs} in the northern \ac{gom} (\citealt{Fennel13,Obenour15,Pauer16}, \citetalias{LehrterIR}).  This area drains a significant portion of the continental United States through the \ac{marb} and is the second largest hypoxic area in the world \citep{Rabalais02}.  Understanding processes that contribute to the frequency and duration of hypoxic events remains a critical research goal for the region, including the application of process-based models to characterize the current knowledge domain.

@@

Standard approaches for uncertainty analysis can be used to begin addressing model complexity issues.  In the most general sense, uncertainty is evaluated relative to the effects of input conditions or the observed data used to calibrate a model, changes in parameter values, or variation in the structural components \citep{Beck87}.  Evaluating parameter uncertainty is by far the most common and simplest means of evaluating model behavior.  Although uncertainty analyses should be integrated throughout model development and application, parameters are more often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is sometimes called inverse modelling where results from sensitivity analyses are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability', where only a subset of parameters can be numerically constrained to the data as compared to the entire parameter set.  Redundancies in parameter effects lead to unidentifiable models where optimal solutions may be empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}.  The concept of identifiable parameter subsets is not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in model development \citep{Fasham06}.  Moreover, the inclusion of sensitivity and identifiability analyses in model tuning will require the adoption of selection rules that are context-dependent given the subsets that are possible with large parameter sets \citep[e.g.,][]{Wagener01, Wagener01b}.

This study describes a parameter sensitivity analysis to evaluate identifiability of parameter subsets for a bio-geo-chemical model of hypoxia for the northern \ac{gom}.  We evaluate a simple \ac{zerod} unit of a larger spatial-temporal model to explore relationships between multiple parameter sets and hypoxia dynamics on the \ac{lcs}.  Specifically, we provide empirical results to support the assumption that models are generally over-parameterized and only a finite and smaller subset of the larger parameter set can be optimized for a given research question or dataset.  We provide explicit guidance for choosing such subsets of the parameter space given constraints on identifiability as directly related to sensitivity analyses.  The objectives are to \begin{inparaenum}[1\upshape)]
\item identify the parameters that have the greatest influence on \ac{do} using local sensitivity analysis,
\item quantify the identifiability of subsets of the total parameter space based on sensitivity,
\item and provide a set of heuristics for choosing parameters based on sensitivity, identifiability, and parameter categories, including extension to other state variables provided by the model.
\end{inparaenum}
A final analysis evaluates identifiability relative to structural uncertainty to provide an example of extending these methods to more complex uncertainty assessments.  The optimum parameter space is defined as the chosen subset that represents the maximum number of identifiable parameters.  Here, `optimum' is both a qualitative description based on a research question or management goal and a quantitative objective based on numerical optimization criteria for fitting model output to a calibration dataset.  These results can be used to refine existing models or guide application of models to novel contexts, such as downscaling or application to new environments.  We conclude with a discussion of the implications for hypoxia formation in coastal regions, including management strategies for nutrient reduction and use of mechanistic models to inform decision-making.

@@

This study evaluated the core unit of a recently developed hydrodynamic and ecological model that describes horizontal and vertical transport and mixing of state variables relevant for hypoxia in the northern \ac{gom}.  The \ac{cgem} includes elements from the Navy Coastal Ocean Model \citep{Martin00} that describes hydrodynamics on the \ac{lcs} and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf (depth $\leq$ 100 m, with additional hybrid layers at deeper depths).  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.  Atmospheric and hydrological boundary conditions described in \citet{Hodur97} and \citet{Lehrter13} are also included in \ac{cgem}.

The core unit of \ac{cgem} is FishTank, a \ac{zerod} model that implements the biogeochemical equations in \citet{Eldridge10} and does not include any form of physical transport (i.e., advection, mixing, or surface flux) nor sediment diagenesis.  Although FishTank was developed for specific application in \ac{cgem}, it can easily be applied to other hydrodynamic grids. Accordingly, the sensitivity and identifiability analyses described below focused exclusively on FishTank and are informative for both the \ac{lcs} gridded model as well as potential applications to different systems.  The FishTank model provides estimates for the 36 state variables described above using a \ac{zerod} parcel that is uniformly mixed as a closed system.  A set of initial conditions is provided on execution of the model that is based on observations of relevant variables obtained from research cruises in the northern \ac{gom} during April, June, and September of 2006 \citep[Table 1 in][]{Murrell14}.

@@

The analysis focused on sensitivity of \ac{do} and other state variables (noted below) in the \ac{zerod} FishTank model to identify parameters that may affect spatial and temporal variation of hypoxia in the larger model.  A local sensitivity analysis was performed by evaluating the change in \ac{do} following perturbation of each parameter from its original value.  The analyses relied exclusively on concepts used in the FME package developed for the R statistical programming language \citep{Soetaert10,RDCT16}. Parameters were individually perturbed by 50\% of the original values and the model was executed to obtain an estimate \ac{do} sensitivity.  For each perturbation, a sensitivity value $S$ was estimated for each time step $i$ given a change for parameter $j$ as:

@@

All parameters for each of the six equation categories (optics, organic matter, phytoplankton, temperature, and zooplankton) that had non-zero $L1$ (suggesting sensitivity) were retained for identifiability analysis.

@@

\noindent where $\gamma$ ranges from one to infinity for perfectly identifiable (orthogonal) or unidentifiable (perfectly collinear) results for parameters in a sensitivity matrix $S$.  The sensitivity functions were supplied as a matrix $\hat{S}$ with rows $i$ and columns $j$ (\cref{sijeqn}) that described deviations of predicted \ac{do} from the default parameter values.  The matrix $\hat{S}$ was first normalized by dividing by the square root of the summed residuals \citep{Omlin01,Soetaert10}.

The collinearity index $\gamma$ provides a measure of the linear dependence between sensitivity functions described above for subsets of parameters. Estimates of $\gamma$ greater than 10-15 suggest parameter sets are poorly identifiable \citep{Brun01,Omlin01}, meaning parameter values that maximize precision on a calibration dataset are inestimable by conventional optimization algorithms given similar effects of the selected parameters on \ac{do}. Greater sensitivity of a state variable to a subset of parameters does not always imply better identifiability if the individual effects are similar.  An intuitive interpretation of $\gamma$ is provided by \citet{Brun01} such that a change in a state variable caused by a change in one parameter can be offset by the fraction $1 - 1/\gamma$ by the remaining parameters.  That is, $\gamma = 10$ suggests the relative change in \ac{do} for an arbitrary parameter in the selected set can be compensated for by 90\% with changes in the other parameters.

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the \Sexpr{length(p1z1pars)} evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., practicioners may only be interested in optics parameters).  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability limits for parameter subsets, and the difficulty of choosing from many combinations.

@@

Parameter sensitivity was evaluated by comparing the results from the default model setup to a more complex setup with alternative switches (\cref{tab:strcs}).  The scenarios used switches for different equations to represent structural relationships of phytoplankton growth patterns with temperature, nutrient uptake, cell quotas, \ac{chla} to carbon ratios, photosynthesis dynamics, and specific growth limits. The default scenario modelled phytoplankton growth with temperature as a sigmoidal function \citep{Eldridge10}, nutrient uptake as Michaelis-Menten kinetics \citep{Dugdale67}, internal cell quotas following \citet{Droop73}, \ac{chla} to carbon ratios as a simple regression \citep{Murrell14}, light-dependence of photosynthesis with photoinhibition \citep{Platt80}, and a specific growth rate following Leibig's law of the minimum.  Conversely, the complex scenario used switches that modelled relationships between phytoplankton growth with temperature using the Arrhenius model \citep{Geider97}, nutrient uptake as proposed by \citet{Lehman75} and modified in \citet{Geider98}, internal cell quotas following \citet{Flynn03}, \ac{chla} to carbon ratios following \citet{Cloern95}, light-dependence of photosynthesis that is nutrient dependent \citep{Flynn03}, and a specific growth rate that is nutrient dependent.  Complete details are provided as supplementary material to \citetalias{LehrterIR}.

@@

Local sensitivity analyses for the additional state variables (ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus) had similar results as \ac{do} with some exceptions (\cref{fig:sensalltile,tab:nh4sens,tab:chlsens,tab:irrsens,tab:no3sens,tab:om1sens,tab:om2sens,tab:po4sens}).  All additional variables were sensitive to the same parameters (\Sexpr{locO2$tot} of \Sexpr{length(p1z1pars)} evaluated), although average sensitivity differed between variables.  Average $L1$ ranged from \Sexpr{scinot(aveall[2])} for \Sexpr{names(aveall)[2]} (\cref{tab:irrsens}) to \Sexpr{scinot(aveall[length(aveall)])} for \Sexpr{names(aveall)[length(aveall)]} (\cref{tab:om2sens}).  All average sensitivity values for the state variables were higher than the average for \ac{do} ($L1$ = \Sexpr{scinot(aveall[1])}).  For each variable, $L1$ ranged from \Sexpr{locNH4$sensrng[1, 2]} (\Sexpr{locNH4$sensrng[1, 1]}) to \Sexpr{locNH4$sensrng[2, 2]} (\Sexpr{locNH4$sensrng[2, 1]}) for ammonium (\cref{tab:nh4sens}), \Sexpr{locChla_mg_tot$sensrng[1, 2]} (\Sexpr{locChla_mg_tot$sensrng[1, 1]}) to \Sexpr{locChla_mg_tot$sensrng[2, 2]} (\Sexpr{locChla_mg_tot$sensrng[2, 1]}) for \ac{chla} (\cref{tab:chlsens}), \Sexpr{locirradiance$sensrng[1, 2]} (\Sexpr{locirradiance$sensrng[1, 1]}) to \Sexpr{locirradiance$sensrng[2, 2]} (\Sexpr{locirradiance$sensrng[2, 1]}) for irradiance (\cref{tab:irrsens}), \Sexpr{locNO3$sensrng[1, 2]} (\Sexpr{locNO3$sensrng[1, 1]}) to \Sexpr{locNO3$sensrng[2, 2]} (\Sexpr{locNO3$sensrng[2, 1]}) for nitrate (\cref{tab:no3sens}), \Sexpr{locOM1$sensrng[1, 2]} (\Sexpr{locOM1$sensrng[1, 1]}) to \Sexpr{locOM1$sensrng[2, 2]} (\Sexpr{locOM1$sensrng[2, 1]}) for \ac{pom} (\cref{tab:om1sens}),  \Sexpr{locOM2$sensrng[1, 2]} (\Sexpr{locOM2$sensrng[1, 1]}) to \Sexpr{locOM2$sensrng[2, 2]} (\Sexpr{locOM2$sensrng[2, 1]}) for \ac{dom} (\cref{tab:om2sens}), and \Sexpr{locPO4$sensrng[1, 2]} (\Sexpr{locPO4$sensrng[1, 1]}) to \Sexpr{locPO4$sensrng[2, 2]} (\Sexpr{locPO4$sensrng[2, 1]}) for phosphate (\cref{tab:po4sens}).  For the parameter categories, ammonium was most sensitive to \Sexpr{names(maxcat$NH4)} parameters (average $L1$ = \Sexpr{maxcat$NH4} across all parameters in the category), \ac{chla} to \Sexpr{names(maxcat$Chla_mg_tot)} ($L1$ = \Sexpr{maxcat$Chla_mg_tot}), irradiance to \Sexpr{names(maxcat$irradiance)} ($L1$ = \Sexpr{maxcat$irradiance}), nitrate to \Sexpr{names(maxcat$NO3)} ($L1$ = \Sexpr{maxcat$NO3}), \ac{pom} to \Sexpr{names(maxcat$OM1)} ($L1$ = \Sexpr{maxcat$OM1}), \ac{dom} to \Sexpr{names(maxcat$OM2)} ($L1$ = \Sexpr{maxcat$OM2}), and phosphate to \Sexpr{names(maxcat$PO4)} ($L1$ = \Sexpr{maxcat$PO4}).  Finally, average sensititivy between parameter categories independent of the state variables ranged from \Sexpr{avecat[1]} for \Sexpr{names(avecat)[1]} (average $L1$ across all variables) to \Sexpr{avecat[length(avecat)]} for \Sexpr{names(avecat)[length(avecat)]}.

@@

# final table formatting
parms <- totab$Parameter
cats <- as.character(totab$Category)
cap_val<- 'Parameter identifiability (as $\\gamma$, \\cref{gameq}) by category for relevant state variables.  Selections followed the first heuristic where parameters were selected within categories from most to least sensitive until $\\gamma > 15$.  Rank describes the relative parameter sensitivity in each category for each state variable. Duplicate parameters and ranks in the first two columns apply only to $\\gamma$ values in the same row (i.e., the top-ranked parameter could vary for each variable).'
colnms <- c('Rank', 'Ammonium', 'Chl-\\textit{a}', 'O$_2$', 'Irradiance', 'Nitrate', 'OM1', 'OM2', 'Phosphate')

>>>>

\begin{abstract}
\noindent Bio-geo-chemical models are useful tools in environmental sciences that can guide management and policy-making. Consequently, significant time and resources are spent developing these models in system-specific contexts. The optimization of model parameters to maximize precision, including transferability of these models to different systems, are fundamental concerns in the development and application of these tools. This study provides a context for understanding quantitative limitations of coupled hydrodynamic-ecological models by evaluating parameter sensitivity and identifability of a \ac{zerod} unit of a larger spatio-temporal model of hypoxia on the \acl{lcs} of Gulf of Mexico.  The analysis provides a contrast of numeric and ecological certainty in parameter subsets using a systematic framework to infer larger trends in dissolved oxygen dynamics over time, having implications for understanding factors that contribute to environmental conditions that are detrimental to aquatic resources.  In particular, we focus on issues of parameter identifiability using local sensitivity analyses to provide quantitative descriptions of numerical constraints on model precision.  We argue that quantitative and ecological certainty in model calibration are often at odds and practitioners must choose explicit model components to optimize given tradeoffs between the two. We further conclude that numerically optimal parameter sets for models of hypoxia are often small subsets of the complete parameter set because of redundancies in the unique effects of parameter perturbations on model output.  As a result, we demonstrate that use of a model for inference into ecological mechanisms of observed or predicted changes in hypoxic condition can be potentially misguided in the absence of quantitative descriptions of identifiability.  Although these concerns have been expressed in the literature, they are rarely explicitly addressed or included in model evaluations.  In addition to immediate implications for regional models, we provide a framework for describing the effects of parameter uncertainty and identifiability that can be applied to similar models to better inform environmental management.
\end{abstract}
\acresetall

@@

Hypoxia formation in bottom waters of coastal oceans occurs primarily from excess nutrient inputs from land-based sources \citep{Justic87,Diaz95,Howarth96}.  These events are detrimental to aquatic organisms and have significant negative effects on economic resources derived from coastal ecosystems \citep{Lipton03,Diaz11}.  An understanding of the biological, physical, and chemical processes that contribute to the growth of hypoxic areas is a critical concern for mitigating and preventing these negative impacts.  Numerical ecosystem models are important tools that synthesize knowledge of ecosystem processes that contribute to hypoxia formation and for predicting the effects of proposed management activities or future scenarios \citep{Scavia04,Hagy07,Pauer16}.  Unlike statistical models with more generic structures, simulation and process-based models include explicit descriptions of relevant processes that are constrained by empirical or observational data relevant to the system of interest \citep[e.g.][]{Omlin01b,Eldridge10}.  These models are often coupled with hydrodynamic grids to provide spatially-explicit representations of patterns in three dimensions \citep{Warner05,Zhao10,Ganju16}. Combined hydrodynamic and bio-geo-chemical models have been developed specifically to describe hypoxic conditions on the \ac{lcs} in the northern \ac{gom} (\citealt{Fennel13,Obenour15,Pauer16}, \citetalias{LehrterIR}).  This area drains a significant portion of the continental United States through the \ac{marb} and is the second largest hypoxic area in the world \citep{Rabalais02}.  Understanding processes that contribute to the frequency and duration of hypoxic events remains a critical research goal for the region, including the application of process-based models to characterize the current knowledge domain.

@@

Standard approaches for uncertainty analysis can be used to begin addressing model complexity issues.  In the most general sense, uncertainty is evaluated relative to the effects of input conditions or the observed data used to calibrate a model, changes in parameter values, or variation in the structural components (i.e., observational, parameter, or structural uncertainty) \citep{Beck87}.  Evaluating parameter uncertainty is by far the most common and simplest means of evaluating model behavior.  Although uncertainty analyses should be integrated throughout model development and application, parameters are more often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is sometimes called inverse modelling where results from sensitivity analyses are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability', where only a subset of parameters can be numerically constrained to the data as compared to the entire parameter set.  Redundancies in parameter effects lead to unidentifiable models where optimal solutions may be empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}. An unidentifiable parameter or parameter set has effects on model output that can be undone or compensated for by alteration of other parameters.  The concept of identifiable parameter subsets is not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in model development \citep{Fasham06}.  Moreover, the inclusion of sensitivity and identifiability analyses in model tuning will require the adoption of conservative selection rules for parameters to calibrate given the number of unique combinations of parameter subsets for most models \citep[e.g.,][]{Wagener01, Wagener01b}.

This study describes a parameter sensitivity analysis to evaluate identifiability of parameter subsets for a bio-geo-chemical model of hypoxia for the northern \ac{gom}.  We evaluate a simple \ac{zerod} unit of a larger spatial-temporal model to explore relationships between multiple parameter sets and hypoxia dynamics on the \ac{lcs}.  Specifically, we provide empirical results to support the assumption that models are generally over-parameterized and only a finite and smaller subset of the larger parameter set can be optimized for a given research question or dataset.  We provide explicit guidance for choosing such subsets of the parameter space given constraints on identifiability as directly related to sensitivity analyses.  The objectives are to \begin{inparaenum}[1\upshape)]
\item identify the parameters that have the greatest influence on \ac{do} using local sensitivity analysis,
\item quantify the identifiability of subsets of the total parameter space based on sensitivity,
\item and provide a set of heuristics for choosing parameters based on sensitivity, identifiability, and parameter categories.
\end{inparaenum}
These principles were also applied to other state variables predicted by the model (ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus). A final analysis evaluated identifiability relative to structural uncertainty to provide an example of extending these methods to more complex uncertainty assessments.  Throughout, the optimum parameter set is defined as the chosen subset that represents the maximum number of identifiable parameters. `Optimum' is both a qualitative description based on a research question or management goal and a quantitative objective based on numerical optimization criteria for fitting model output to a calibration dataset.  These results can be used to refine existing models or guide application of models to novel contexts, such as downscaling or application to new environments.  We conclude with a discussion of the implications for hypoxia formation in coastal regions, including management strategies for nutrient reduction and use of mechanistic models to inform decision-making.

@@

This study evaluated the core unit of a recently developed hydrodynamic and ecological model that describes horizontal and vertical transport and mixing of state variables relevant for hypoxia in the northern \ac{gom}.  The \ac{cgem} includes elements from the Navy Coastal Ocean Model \citep{Martin00} for hydrodynamics on the \ac{lcs} and a biogeochemical model with multiple plankton groups, water-column metabolism, and sediment diagenesis \citep{Eldridge10}.  The hydrodynamic component of \ac{cgem} provides a spatially-explicit description of hypoxia using an orthogonal grid with an approximate horizontal resolution of 1.9 km$^2$ and twenty equally-spaced vertical sigma layers on the shelf (depth $\leq$ 100 m, with additional hybrid layers at deeper depths).  The biogeochemical component includes equations for 36 state variables including six phytoplankton groups (with nitrogen and phosophorus quotas for each), two zooplankton groups, nitrate, ammonium, phosphate, dissolved inorganic carbon, oxygen, silica, and multiple variables for dissolved and particulate organic matter from different sources.  Atmospheric and hydrological boundary conditions described in \citet{Hodur97} and \citet{Lehrter13} are also included in \ac{cgem}.

The core unit of \ac{cgem} is FishTank, a \ac{zerod} model that implements the biogeochemical equations in \citet{Eldridge10} and does not include any form of physical transport (i.e., advection, mixing, or surface flux) nor sediment diagenesis.  Although FishTank was developed for specific application in \ac{cgem}, it can easily be applied to other hydrodynamic grids. Accordingly, the sensitivity and identifiability analyses described below are informative for both the \ac{lcs} gridded model as well as potential applications to different systems.  The FishTank model provides estimates for the 36 state variables described above using a \ac{zerod} parcel that is uniformly mixed as a closed system.  A set of initial conditions is provided on execution of the model that is based on observations of relevant variables obtained from research cruises in the northern \ac{gom} during April, June, and September of 2006 \citep[Table 1 in][]{Murrell14}.

@@

The analysis focused on sensitivity of \ac{do} and other state variables (noted below) in the \ac{zerod} FishTank model to identify parameters that may affect spatial and temporal variation of hypoxia in the larger model.  A local sensitivity analysis was performed by evaluating the change in \ac{do} following perturbation of each parameter from its original value.  The analyses relied exclusively on concepts used in the FME package developed for the R statistical programming language \citep{Soetaert10,RDCT16}. Parameters were individually perturbed by 50\% of the original values and the model was executed to obtain an estimate of \ac{do} sensitivity.  For each perturbation, a sensitivity value $S$ was estimated for each time step $i$ given a change for parameter $j$ as:

@@

The $L1$ value for each parameter was used as the primary measure of sensitivity for the state variables. All parameters for each of the six equation categories (optics, organic matter, phytoplankton, temperature, and zooplankton) that had non-zero $L1$ (suggesting sensitivity) were retained for identifiability analysis.

@@

\noindent where $\gamma$ ranges from one to infinity for perfectly identifiable (orthogonal) or unidentifiable (perfectly collinear) results for parameters in a sensitivity matrix $S$.  The sensitivity functions were supplied as a matrix $\hat{S}$ with rows $i$ and columns $j$ (\cref{sijeqn}) that described deviations of predicted \ac{do} from the default parameter values.  Thus, $\gamma$ can be estimated for any subset of parameter combinations using the change in model output for perturbations of individual parameters. Sensitivity matrices were first normalized by dividing by the square root of the summed residuals \citep{Omlin01,Soetaert10}.

The collinearity index $\gamma$ provides a measure of the linear dependence between sensitivity functions (i.e., $S_i$ for $j$ parameters) described above for subsets of parameters. Estimates of $\gamma$ greater than 10-15 suggest parameter sets are poorly identifiable \citep{Brun01,Omlin01}, meaning parameter values that maximize precision on a calibration dataset are inestimable by conventional optimization algorithms given similar effects of the selected parameters on the estimated state variable. Greater sensitivity of a state variable to a subset of parameters does not always imply better identifiability if the effects of individual parameters are similar.  An intuitive interpretation of $\gamma$ is provided by \citet{Brun01} such that a change in a state variable caused by a change in one parameter can be offset by the fraction $1 - 1/\gamma$ by the remaining parameters.  That is, $\gamma = 10$ suggests the relative change in \ac{do} for an arbitrary parameter in the selected set can be compensated for by 90\% with changes in the other parameters.

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the \Sexpr{length(p1z1pars)} evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., practitioners may only be interested in optics parameters).  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability limits for parameter subsets, and the difficulty of choosing from many combinations.

@@

Parameter sensitivity was evaluated by comparing the results from the default model setup to a more complex setup with alternative switches (\cref{tab:strcs}).  The scenarios used switches for different equations to represent structural relationships of phytoplankton growth patterns with temperature, nutrient uptake, cell quotas, \ac{chla} to carbon ratios, photosynthesis dynamics, and specific growth limits. The default scenario modelled phytoplankton growth with temperature as a sigmoidal function \citep{Eldridge10}, nutrient uptake as Michaelis-Menten kinetics \citep{Dugdale67}, internal cell quotas following \citet{Droop73}, \ac{chla} to carbon ratios as a simple regression \citep{Murrell14}, light-dependence of photosynthesis with photoinhibition \citep{Platt80}, and a specific growth rate following Leibig's law of the minimum.  Conversely, the complex scenario used switches that modelled relationships between phytoplankton growth with temperature using the Arrhenius model \citep{Geider97}, nutrient uptake as proposed by \citet{Lehman75} and modified in \citet{Geider98}, internal cell quotas following \citet{Flynn03}, \ac{chla} to carbon ratios following \citet{Cloern95}, light-dependence of photosynthesis that is nutrient dependent \citep{Flynn03}, and a specific growth rate that is nutrient dependent.  Complete details of model switches are provided as supplementary material to \citetalias{LehrterIR}.

@@

Local sensitivity analyses for the additional state variables (ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus) had similar results as \ac{do} with some exceptions (\cref{fig:sensalltile,tab:nh4sens,tab:chlsens,tab:irrsens,tab:no3sens,tab:om1sens,tab:om2sens,tab:po4sens}).  All variables were sensitive to the same parameters as \ac{do} (\Sexpr{locO2$tot} of \Sexpr{length(p1z1pars)} evaluated), although average sensitivity differed between variables.  Average $L1$ ranged from \Sexpr{scinot(aveall[2])} for \Sexpr{names(aveall)[2]} (\cref{tab:irrsens}) to \Sexpr{scinot(aveall[length(aveall)])} for \Sexpr{names(aveall)[length(aveall)]} (\cref{tab:om2sens}).  All average sensitivity values for the state variables were higher than the average for \ac{do} ($L1$ = \Sexpr{scinot(aveall[1])}).  For each variable, $L1$ ranged from \Sexpr{locNH4$sensrng[1, 2]} (\Sexpr{locNH4$sensrng[1, 1]}) to \Sexpr{locNH4$sensrng[2, 2]} (\Sexpr{locNH4$sensrng[2, 1]}) for ammonium (\cref{tab:nh4sens}), \Sexpr{locChla_mg_tot$sensrng[1, 2]} (\Sexpr{locChla_mg_tot$sensrng[1, 1]}) to \Sexpr{locChla_mg_tot$sensrng[2, 2]} (\Sexpr{locChla_mg_tot$sensrng[2, 1]}) for \ac{chla} (\cref{tab:chlsens}), \Sexpr{locirradiance$sensrng[1, 2]} (\Sexpr{locirradiance$sensrng[1, 1]}) to \Sexpr{locirradiance$sensrng[2, 2]} (\Sexpr{locirradiance$sensrng[2, 1]}) for irradiance (\cref{tab:irrsens}), \Sexpr{locNO3$sensrng[1, 2]} (\Sexpr{locNO3$sensrng[1, 1]}) to \Sexpr{locNO3$sensrng[2, 2]} (\Sexpr{locNO3$sensrng[2, 1]}) for nitrate (\cref{tab:no3sens}), \Sexpr{locOM1$sensrng[1, 2]} (\Sexpr{locOM1$sensrng[1, 1]}) to \Sexpr{locOM1$sensrng[2, 2]} (\Sexpr{locOM1$sensrng[2, 1]}) for \ac{pom} (\cref{tab:om1sens}),  \Sexpr{locOM2$sensrng[1, 2]} (\Sexpr{locOM2$sensrng[1, 1]}) to \Sexpr{locOM2$sensrng[2, 2]} (\Sexpr{locOM2$sensrng[2, 1]}) for \ac{dom} (\cref{tab:om2sens}), and \Sexpr{locPO4$sensrng[1, 2]} (\Sexpr{locPO4$sensrng[1, 1]}) to \Sexpr{locPO4$sensrng[2, 2]} (\Sexpr{locPO4$sensrng[2, 1]}) for phosphate (\cref{tab:po4sens}).  For the parameter categories, ammonium was most sensitive to \Sexpr{names(maxcat$NH4)} parameters (average $L1$ = \Sexpr{maxcat$NH4} across all parameters in the category), \ac{chla} to \Sexpr{names(maxcat$Chla_mg_tot)} ($L1$ = \Sexpr{maxcat$Chla_mg_tot}), irradiance to \Sexpr{names(maxcat$irradiance)} ($L1$ = \Sexpr{maxcat$irradiance}), nitrate to \Sexpr{names(maxcat$NO3)} ($L1$ = \Sexpr{maxcat$NO3}), \ac{pom} to \Sexpr{names(maxcat$OM1)} ($L1$ = \Sexpr{maxcat$OM1}), \ac{dom} to \Sexpr{names(maxcat$OM2)} ($L1$ = \Sexpr{maxcat$OM2}), and phosphate to \Sexpr{names(maxcat$PO4)} ($L1$ = \Sexpr{maxcat$PO4}).  Finally, average sensitivity between parameter categories independent of the state variables ranged from \Sexpr{avecat[1]} for \Sexpr{names(avecat)[1]} (average $L1$ across all variables) to \Sexpr{avecat[length(avecat)]} for \Sexpr{names(avecat)[length(avecat)]}.

@@

# final table formatting
parms <- totab$Parameter
cats <- as.character(totab$Category)
cap_val<- 'Parameter identifiability (as $\\gamma$, \\cref{gameq}) by category for relevant state variables.  Selections followed the first heuristic where parameters were selected within categories from most to least sensitive until $\\gamma > 15$.  Rank describes the relative parameter sensitivity in each category for each state variable. Duplicate parameters and ranks in the first two columns apply only to $\\gamma$ values in the same row (i.e., parameter ranks vary for each variable).'
colnms <- c('Rank', 'Ammonium', 'Chl-\\textit{a}', 'O$_2$', 'Irradiance', 'Nitrate', 'OM1', 'OM2', 'Phosphate')
